\documentclass[8pt]{beamer}
\setbeamertemplate{navigation symbols}{}  % remove navigation symbols

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amssymb,amsthm,amsmath}

\hypersetup{colorlinks,linkcolor=,urlcolor=magenta}

\lstdefinelanguage{Julia}%
  {morekeywords={abstract,begin,break,case,catch,const,continue,do,else,elseif,%
      end,export,false,for,function,immutable,import,importall,if,in,%
      macro,module,otherwise,quote,return,struct,switch,true,try,typealias,%
      using,while,Union,Any,Number,AbstractArray,T,U},%
   sensitive=true,%
   alsoother={\$},%
   morecomment=[l]\#,%
   morecomment=[n]{\#=}{=\#},%
   morestring=[s]{"}{"},%
   morestring=[m]{'}{'},%
   basicstyle = \ttfamily,
   keywordstyle = \color{purple},
   keywordstyle = [2]{\color{violet}},
   morekeywords = [2]{trait,implement}
}[keywords,comments,strings]%


\lstset{
    language = Julia,
    escapeinside={/*}{*/},
    basicstyle=\normalfont
}

\lstdefinelanguage{sml}{
  morekeywords={exn},
  keywordstyle = \color{blue}
}

% code inline
\newrobustcmd{\sml}[2][]{{\sloppy
\ifmmode
\text{\lstinline[language=sml,#1]`#2`}
\else{\lstinline[language=sml,#1]`#2`}%
\fi}}

\newrobustcmd{\julia}[2][]{{\sloppy
\ifmmode
\text{\lstinline[language=Julia,#1]`#2`}
\else{\lstinline[language=Julia,#1]`#2`}%
\fi}}

\newcommand{\yikes}{\textbf{\color{red} (!) }}

\begin{document}

\begin{frame}
  \frametitle{Type Unions}

  Terminology: \emph{types} are prescriptive. Some examples:
  \begin{align*}
    \sml{unit}  &:= \sml{()} \\
    \sml{bool}  &:= \sml{false} \mid \sml{true} \\
    \sml{order} &:= \sml{LESS} \mid \sml{EQUAL} \mid \sml{GREATER} \\
    \sml{nat}   &:= \sml{zero} \mid \sml{succ(nat)}
  \end{align*}

  \textit{Product types} (Cartesian product, tuples): $\tau_1 \times \tau_2$.

  \noindent\rule{\linewidth}{0.4pt}

  What about ``combining'' two types? One idea: $\julia{Union\{}\tau_1,\tau_2\julia{\}}$. However:
  \begin{align*}
    &\julia{const MightFail\{T\} = Union\{T,String\}} \\
    &\julia{const IntMightFail = MightFail\{Int\}} &&\julia{Union\{Int,String\}} \\
    &\julia{const StringMightFail = MightFail\{String\}} &&\julia{String} \yikes
  \end{align*}

  Alternative: \textit{sum types} (``tagged disjoint unions''): $\tau_1 + \tau_2$.
  \begin{align*}
    &\julia{const MightFail\{T\} = Ok\{T\} | Error\{String\}}
  \end{align*}

  Benefits:
  \begin{itemize}
    \item More modular -- cases are necessarily disjoint.
    \item Easier to develop efficient data structures, since sums are ``concrete''.
    \item Guaranteed \emph{case exhaustivity} checking! Easy to refactor code.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Dynamic Typing and Dynamic Dispatch}

  In PL theory, types are checked without executing code (i.e. at ``compile-time''). Therefore, ``types'' in Julia aren't types, formally. Instead, Julia has one type -- \julia{Any}, an \emph{extensible sum}\footnote{\href{http://www.cs.cmu.edu/~rwh/pfpl/}{Practical Foundations for Programming Languages, Second Edition (Harper)}: Chapter 33} -- where each ``type'' is a \emph{case}, tagging a value.

  \begin{align*}
    &\julia{Any} := \julia{Int(int)} \mid \julia{Float(float)} \mid \julia{String(string)} \mid \dots
  \end{align*}

  \begin{itemize}
    \item All Julia values (of type \julia{Any}) are ``tagged''. Therefore, calling a function requires a case analysis (``method table lookup''), incurring a runtime performance penalty. \yikes This explains why dynamic dispatch is avoided in critical-path code!
    \item Similarly, since each Julia ``function'' is a mutable registry covering domain \julia{Any}, type piracy is inherently possible.
  \end{itemize}

  \noindent\rule{\linewidth}{0.4pt}

  Conclusions:
  \begin{itemize}
    \item This is \emph{less general} than having multiple types! There's no way to guarantee a given input has a given shape; instead, programmers must rely on knowledge of how the compiler optimizes in order to write performant code.
    \item However, \julia{Any} can coexist with other types! For example, ML\footnote{\href{https://smlfamily.github.io/sml97-defn.pdf}{The Definition of Standard ML (Milner, Tofte, Harper, MacQueen)}}-style languages support generative types alongside \sml{exn}, an extensible sum type, like Julia's \julia{Any}.
    \item Otherwise, gradual types%
    \footnote{\href{https://wphomes.soic.indiana.edu/jsiek/what-is-gradual-typing/}{What is Gradual Typing (Siek)}}$^,$%
    \footnote{\href{http://www.ccs.neu.edu/home/amal/papers/gtt.pdf}{Gradual Type Theory (New, Licata, Ahmed)}}
    could be the answer?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Abstract Types}

  What does it mean for \julia{x::T}, with \julia{T} abstract?
  \begin{itemize}
    \item \julia{x::U}, for some concrete \julia{U} with \julia{U <: T}
  \end{itemize}

  What properties should \julia{U} have if \julia{U <: T}? Often, there is an (implicit) collection of functions which should be implemented for \julia{U}. For example:
  \begin{itemize}
    \item If \julia{U <: Number}, then \julia{Base.:+(::U, ::U)::U}, \julia{Base.:*(::U, ::U)::U}, etc. should be defined.
    \item If \julia{U <: AbstractArray\{Elt\}}, then \julia{Base.length(::U)::Int}, \julia{Base.getindex(::U, ::Int)::Elt}, etc. should be defined.
  \end{itemize}

  Therefore, \julia{U <: T} means that \julia{U} implements an interface specified by \julia{T}. In other languages, this feature is often called \emph{\href{http://learnyouahaskell.com/types-and-typeclasses}{typeclasses}} or \emph{\href{https://doc.rust-lang.org/book/ch10-02-traits.html}{traits}}.%
  \footnote{\href{https://github.com/HarrisonGrodin/radical-julia}{\texttt{radical-julia} (Grodin)}}

  \noindent\rule{\linewidth}{0.4pt}

  Benefits:
  \begin{itemize}
    \item Traits can still form a hierarchy. For example, $\julia{trait Ordered\{T\} <: Eq\{T\}}$.
    \item Instances can be automatically derived. For example, $\julia{implement Eq\{Vector\{T\}\} where Eq\{T\}}$.
    \item Traits can \emph{also} form more complex lattices (``multiple inheritance'')!
    \item Similar to dynamic dispatch, traits simulate ``overloading,'' allowing a given function to be implemented differently for various types. However, traits are a zero-cost abstraction\footnote{\href{https://blog.rust-lang.org/2015/05/11/traits.html}{Abstraction without overhead: traits in Rust (Turon)}}, allowing the compiler to generate specialized code \emph{before} runtime, leading to faster execution.
    \item No more runtime ambiguity errors!
  \end{itemize}
\end{frame}

\end{document}
